Approximation as Differentiation: Notes

### 2025-05-27 main/formal-structures.tex
Correction.
Tweaks.

Our approach to a denotational account of \GPS is based on the CHAD (Combinatory Homomorphic Automatic Differentiation) work of Mattias Vákár and others~\cite{vákár22,nunes2023}. Before we go into details, we present a high-level overview of their approach and how we have applied it to \GPS.

@@

Happily, $\FinVect$ and $\LatGal$ both have biproducts, as a consequence of their having products and being enriched in commutative monoids~(\propref{biproducts:from-product-or-coproduct} below). However, neither of them has all small products, without moving to infinite dimensional vector spaces (in the case of $\FinVect$) or complete lattices (in the case of $\LatGal$). Neither of these are ideal. In infinite dimensional vector spaces, dualisation is not involutive and the connection between the forward and reverse derivatives is lost. In complete lattices, we can no longer easily implement the infinite meets and joins required, and we lose the computability of the model. Moreover, in Agda, there are issues with predicativity when considering complete lattices. \todo{be more specific}

Vákár and collaborators overcome this by using different interpretations for forward and backward derivatives. We overcome this by using $\CMon$-enriched presheaves; we turn to this in \secref{higher-order} but first flesh out the first-order setting.

@@

First we introduce the category of lattices and Galois connections, which will serve as our base category for
interpreting forward and backward \GPS. Galois connections are pairs of monotone functions $f: Y \to
X$ and $g: X \to Y$ between posets, where $f$ is the (pointwise) best approximation from below to an inverse
of $g$, and $g$ the best approximation from above to an inverse of $f$. This setup is a nice fit for the
bidirectional nature of (dynamic) program slicing, with $f$ capturing \emph{backward} slicing (producing the
least input slice for a given slice of the output), and $g$ capturing \emph{forward} slicing (producing the
greatest output slice for a given slice of the input).

@@

As sketched in \secref{introduction:galois-slicing}, for \GPS we would like a setting where the approximants
of a point $x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$ representing
the approximation that discards all information about $x$, greatest element $\top$ representing the
approximation that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways to
combine approximations. Thus rather than working directly in $\PosGal$, we consider the following subcategory
instead.

@@

The meet-preserving maps from $X$ to $Y$ themselves form a meet-semilattice. The operation $\meet$ on such
maps is defined pointwise, so that $(f \meet g)(x) = f(x) \meet g(x): X \to Y$, and preserves meets; the
constant map $\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$, which also preserves meets, provides the
unit. Dually the join-preserving maps from $Y$ to $X$ have a join-semilattice structure given pointwise by
$\join$ and the constant homomorphism $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to $\bot_X$. Since these two
constructions come in adjoint pairs, the hom-set $\LatGal(X,Y)$ forms a bounded meet semilattice with unit
$\top_{X,Y}$ given by the Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and meet of Galois
connections $(f \adj g) \meet (f' \adj g') = (f \join f') \adj (g \meet g'): X \to Y$. Thus $\LatGal$ is
enriched in $\SemiLat$, the category of semilattices and semilattice homomorphisms.

>>>>

Our denotational approach to \GPS is based on the CHAD (Combinatory Homomorphic Automatic Differentiation) work of Mattias Vákár and others~\cite{vákár22,nunes2023}. Before we go into details, we present a high-level overview of their approach and how we have applied it to \GPS.

@@

Happily, $\FinVect$ and $\LatGal$ both have biproducts, as a consequence of their having products and being enriched in commutative monoids~(\propref{biproducts:from-product-or-coproduct} below). However, neither of them has all small products, without moving to infinite dimensional vector spaces (in the case of $\FinVect$) or complete lattices (in the case of $\LatGal$). Neither of these is ideal. In infinite dimensional vector spaces, dualisation is not involutive and the connection between the forward and reverse derivatives is lost. In complete lattices, we can no longer easily implement the infinite meets and joins required, and we lose the computability of the model. Moreover, in Agda, there are issues with predicativity when considering complete lattices. \todo{be more specific}

Vákár and collaborators overcome this by using different interpretations for forward and backward derivatives. We overcome this by using $\CMon$-enriched presheaves, as described in \secref{higher-order}; first we present a suitable first-order setting.

@@

We start with the category of lattices and Galois connections, which will serve as our base category for
interpreting forward and backward \GPS. Galois connections are pairs of monotone functions $f: Y \to X$ and
$g: X \to Y$ between posets, where $f$ is the (pointwise) best approximation from below to an inverse of $g$,
and $g$ the best approximation from above to an inverse of $f$. This setup is a nice fit for the bidirectional
nature of (dynamic) program slicing, with $f$ capturing \emph{backward} slicing (producing the least input
slice for a given slice of the output), and $g$ capturing \emph{forward} slicing (producing the greatest
output slice for a given slice of the input).

@@

As sketched in \secref{introduction:galois-slicing}, for \GPS we would like a setting where the approximants
of a point $x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$ representing
the approximation that discards all information about $x$, greatest element $\top$ representing the
approximation that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways to
combine approximations. Thus rather than working directly in $\PosGal$, we will use the following subcategory
instead.

@@

The meet-preserving maps from $X$ to $Y$ themselves form a meet-semilattice. The operation $\meet$ on such
maps is given pointwise, so that $(f \meet g)(x) = f(x) \meet g(x): X \to Y$, and preserves meets; the
constant map $\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$, which also preserves meets, provides the
unit. Dually the join-preserving maps from $Y$ to $X$ have a join-semilattice structure given pointwise by
$\join$ and the constant map $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to $\bot_X$. Since these two
constructions come in adjoint pairs, the hom-set $\LatGal(X,Y)$ forms a bounded meet semilattice with unit
$\top_{X,Y}$ given by the Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and meet of Galois
connections $(f \adj g) \meet (f' \adj g') = (f \join f') \adj (g \meet g'): X \to Y$. Thus $\LatGal$ is
enriched in $\SemiLat$, the category of semilattices and semilattice homomorphisms.
